---
title: "Modeling"
format: html
editor: visual
---

# Introduction

The goal of this analysis is to fit and tune classifiers that predict diabetes status using the survey data. The dataset originates from a large public health survey program, the Behavioral Risk Factor Surveillance System, collected and published by the Centers for Disease Control and Prevention. Since predictors are mixed-type and may relate to diabetes in non-linear ways, we compare two popular tree-based classifiers: a single classification tree model and a random forest, both fit using modeling workflows from the tidymodels package ecosystem. Models are tuned using log-loss so predictions that confidently assign the correct class are rewarded and over-confident wrong predictions are penalized.

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(tidymodels)
library(janitor)

set.seed(987)

#Read in cleaned data from EDA
diabetes <- read_rds("data/diabetes_clean.rds")
```

# Data Split

After loading, we split the data into 70% training and 30% testing to evaluate generalization performance. A seed ensures reproducibility across models.
```{r}
#| message: false
#| warning: false

#Split Data
diab_split <- initial_split(diabetes, prop = 0.7, strata = diabetes_binary)
train_data <- training(diab_split)
test_data <- testing(diab_split)
```

# Classification Tree

A classification tree splits data into regions using predictor values to separate diabetes groups. We tune the cost complexity parameter to balance tree size and generalization, since our EDA showed skew and imbalance that could lead to overfitting without pruning.
```{r}
#| message: false
#| warning: false

#Specifying classification tree model
tree_spec <- decision_tree(
  cost_complexity = tune(),
  min_n = 20
) |>
  set_mode("classification") |>
  set_engine("rpart")

#Workflow (6 predictors)
tree_wf <- workflow() |>
  add_model(tree_spec) |>
  add_formula(
    diabetes_binary ~ bmi + age + high_bp + 
      high_chol + phys_activity + smoker
  )

#5-fold cross-validation (on training set)
tree_folds <- vfold_cv(train_data, v=5, strata = diabetes_binary)

#Tuning grid (for cost complexity)
tree_grid <- grid_regular(
  cost_complexity(),
  levels = 10
)

#Tuning the tree using log-loss
tree_tune <- tune_grid(
  tree_wf,
  resamples = tree_folds,
  grid = tree_grid,
  metrics = metric_set(mn_log_loss),
  control = control_grid(save_pred = TRUE)
)

#Looking at performance 
tree_metrics <- collect_metrics(tree_tune)
tree_metrics

#Selecting lowest log-loss and finalizing work flow
best_tree <- select_best(tree_tune, metric = "mn_log_loss")
best_tree

final_tree_wf <- finalize_workflow(tree_wf, best_tree)

#Fit final tree on training data 
tree_final_fit <- last_fit(final_tree_wf, diab_split)

#Test-set performance
tree_final_fit |> collect_metrics()

#Compute log-loss on test set for the final tree
tree_logloss <- tree_final_fit |>
  collect_predictions() |>
  mn_log_loss(truth = diabetes_binary,
              .pred_Yes,
              event_level = "second")

tree_logloss
```
Test log-loss matches training (≈ 0.35), so the pruned tree generalized well without obvious overfit. We keep this as our best single-tree model and move to random forest tuning.

# Random Forest

A random forest extends the idea of a single classification tree by fitting many trees on bootstrap samples of the training data and averaging their predictions. At each split, a random subset of predictors is considered, which decorrelates the trees and reduces variance relative to a single tree. This makes random forests less interpretable but typically more accurate and stable. Here, we tune the mtry parameter, which controls how many predictors are available at each split, using 5-fold cross-validation and log-loss on the training data.
```{r}
#| message: false
#| warning: false

#Specifying Random Forest Model
rf_spec <- rand_forest(
  mtry  = tune(),
  trees = 100,
  min_n = 20
) |>
  set_mode("classification") |>
  set_engine("ranger", importance = "impurity")

#Workflow (same 6 predictors as Classification Tree)
rf_wf <- workflow() |>
  add_model(rf_spec) |>
  add_formula(
    diabetes_binary ~ bmi + age + high_bp +
      high_chol + phys_activity + smoker
  )

#5-fold cross-validation (on training set)
rf_folds <- vfold_cv(train_data, v = 5, strata = diabetes_binary)

#Tuning grid for mtry
rf_grid <- tibble(mtry = 2:6)

#Tuning random forest using log-loss
rf_tune <- tune_grid(
  rf_wf,
  resamples = rf_folds,
  grid = rf_grid,
  metrics = metric_set(mn_log_loss),
  control = control_grid(save_pred = TRUE)
)

#Look at performance (mtry values)
rf_metrics <- collect_metrics(rf_tune)
rf_metrics

#Selecting lowest log-loss (best mtry) and finalize workflow
rf_best <- select_best(rf_tune, metric = "mn_log_loss")
rf_best

rf_final_wf <- finalize_workflow(rf_wf, rf_best)

#Fit final random forest on training/test split
rf_final_fit <- last_fit(rf_final_wf, diab_split)

#Test-set performance
rf_final_fit |> collect_metrics()

#Compute log-loss on test set for the final random forest
rf_logloss <- rf_final_fit |>
  collect_predictions() |>
  mn_log_loss(truth = diabetes_binary,
              .pred_Yes,
              event_level = "second")

rf_logloss
```

# Final Model Selection
We compare our best pruned classification tree and tuned random forest, selecting the strongest test-set performer as our final mode.
```{r}
#Collect the test-set metrics into objects
tree_test_metrics <- tree_final_fit |> collect_metrics()
rf_test_metrics   <- rf_final_fit   |> collect_metrics()

#Extract test-set log-loss values too
tree_log <- tree_final_fit |> collect_predictions() |> mn_log_loss(diabetes_binary, .pred_Yes, event_level = "second")
rf_log   <- rf_final_fit   |> collect_predictions() |> mn_log_loss(diabetes_binary, .pred_Yes, event_level = "second")

#Putting metrics into a single table
compare <- bind_rows(
  tree_test_metrics |> mutate(model = "Classification Tree"),
  rf_test_metrics   |> mutate(model = "Random Forest")
) |>
  select(model, .metric, .estimate)

#Adding log-loss (not part of default `collect_metrics()` output)
compare <- bind_rows(
  compare,
  tibble(model = "Classification Tree", .metric = "mn_log_loss", .estimate = tree_log$.estimate),
  tibble(model = "Random Forest",       .metric = "mn_log_loss", .estimate = rf_log$.estimate)
)

compare

compare |> knitr::kable(digits = 3)
```
The Random Forest model is selected as the overall winner due to superior ROC AUC and lower log-loss with comparable accuracy.